---
title: "vaccineff"
bibliography: references.bib
link-citations: true
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vaccineff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(vaccineff)
```

## Usage

Vaccines are created to offer protection against diseases that affect human health. Quantifying how well vaccines work in controlled environments and in real-life settings remains a challenge for scientists. Estimating vaccine effectiveness ($VE$) is a key task once a vaccine is available as a control measure within a population, such as during the middle phase of an epidemic like Ebola or COVID-19, and also in the evaluation of regular vaccination programs, such as childhood vaccines.

`vaccineff` provides tools to estimate $VE$ under different study designs [@bookvaccine]. The package provides a set of features for preparing the data, estimating crude and adjusted effectiveness, controlling for potential confounders such as age and assessing the performance of  the models used to approximate $VE$.

## Who are the users / potential users?

`vaccineff` is useful for local, national, and international health agencies looking for a quick implementation to estimate $VE$ based on their available data. It also provides insights to researchers, data analysts, and epidemiology students on how to approach $VE$ using different methods. We believe that `vaccineff` would be specially useful for users without advanced training in statistical methods.


## What is vaccine effectiveness?

In contrast with vaccine efficacy, which is the percentage reduction of disease incidence in a vaccinated group compared with an unvaccinated group under ideal conditions, $VE$ is the percentage reduction of disease incidence in a vaccinated group compared with an unvaccinated group under routine conditions. The reduction attributable to vaccination is usually assessed from data collected in observational studies [@bookvaccine]. Evaluating the effectiveness of vaccines in the field is an important aspect of monitoring immunization programs. 

## For which designs is this package?

`vaccineff` is a package designed to be used for any infectious disease for which a vaccine strategy has been implemented. This current version only allows measuring $VE$ for cohort study designs. Future version will include other designs such as test-negative/case-control studies, and the screening method [@torvaldsen2002observational]. For more information, see the vignette [Other designs](https://epiverse-trace.github.io/vaccineff/articles/other_designs.html).

### Cohort Design

In the cohort design, $VE$ is estimated using the Hazard Ratios ($HR$) between vaccinated and unvaccinated populations,

$$VE = (1-HR(t))\times100.$$

The $HR$ is estimated using the Cox Proportional Hazards model. In particular, we use the vaccine status of the individuals as the only covariate in the regression. Other confounders can be included as matching arguments to adjust for observational bias. The proportional hazards hypothesis is checked using the Schoenfeld test. A visual check is also provided using the log-log representation of the Survival Probability. If the hypothesis is not satisfied, it is recommended to stratify the population into smaller groups using the confounding variables.

## What type of data is needed to use the package?

This package is designed to be used with vaccination data sets with the following structure.

### Data for Cohort design

Data should be disaggregated at the individual level to track vaccinated and unvaccinated populations over time. The dataset must include the following information:

- Date(s) of vaccination for each individual: The package allows for multiple doses per individual and estimates the immunization date using delay times of outcomes and the timing of vaccine administration.

- Date(s) of outcome(s): The package estimates vaccine effectiveness against various outcomes.

- Date(s) of right censoring: The package allows for the inclusion of information on dates of events that constitute right censoring.

- Individuals' demographic information (e.g., sex, age group, health status): These can be used as confounding variables to match the population and reduce observational bias.

An example dataset for a cohort design is included, with information on vaccination dates and biological details per dose, as well as relevant demographic information. The level of data aggregation is tailored to the characteristics and needs of the study case. To load this dataset, run the following code

```{r cohortdata, include = TRUE, echo = TRUE}
# Load example data
data("cohortdata")
head(cohortdata)
```

## Modeling vaccine effectiveness

### VE for Cohort design

The current release of the package bases the estimation of $VE$ in the cohort design on the assumption of proportional hazards between vaccinated and unvaccinated populations. The $HR$ is estimated using the Cox proportional hazards model implemented in the R package `{survival}`.

The integrated dataset `cohortdata` serves as a minimal example of the package's input. The data is accessed using `data("cohortdata")`.


`vaccineff` has three main functions:
1. `make_vaccineff_data`: This function returns an S3 object of the class `vaccineff_data` with the relevant information for the study. This function also allows to create a matched cohort to control for confounding variables by setting `match = TRUE` and passing the corresponding `exact` and `nearest` arguments.
`make_vaccineff_data`  supports the method `summary()` to check the characteristics of the cohort, the matching balance and the sizes of matched, excluded, and removed populations.

2.  `plot_coverage`: This function returns a plot of the vaccine coverage or the cumulative coverage. If the population is matched, the plot also includes the resulting count of doses after matching. 

3. `effectiveness`: This function provides methods for estimating VE using the $HR$. A summary of the estimation can be obtained using `summary()` and a graphical representation of the methodology is generated by `plot().`

```{r getstarted, include = TRUE, echo = TRUE}
# Create `vaccineff_data`
vaccineff_data <- make_vaccineff_data(
  data_set = cohortdata,
  outcome_date_col = "death_date",
  censoring_date_col = "death_other_causes",
  vacc_date_col = "vaccine_date_2",
  vaccinated_status = "v",
  unvaccinated_status = "u",
  immunization_delay = 15,
  end_cohort = as.Date("2044-12-31"),
  match = TRUE,
  exact = c("age", "sex"),
  nearest = NULL
)

# Print summary of vaccineff data object
summary(vaccineff_data)

# Plot the vaccine coverage of the total population
plot_coverage(vaccineff_data)

# Estimate the Vaccine Effectiveness at 90 days
ve90 <- effectiveness(vaccineff_data, at = 90)

# Print summary of VE
summary(ve90)

# Loglog plot to check proportional hazards
plot(ve90, type = "loglog")
```

For details on the estimation of VE in cohort studies see the vignette [Introduction to cohort design with vaccineff](https://epiverse-trace.github.io/vaccineff/articles/cohort_design.html)

## Key references
